Deciding Bitvector Termination with Program Synthesis

### 2014-07-04 papers/termination/paper.tex
isolated loop (non-)termination

\section{Motivational Examples} \label{sec:motivation}
Figure~\ref{fig:handletable} illustrates the most common assumptions made by existent termination analyses:
\begin{itemize}
\item[(i)] programs are linear.
\item[(ii)] terminating programs have linear termination arguments.
\item[(iii)] bit-vectors semantics and mathematical integers semantics are equivalent.
\item[(iv)] floats semantics and mathematical reals semantics are equivalent.
\end{itemize}

@@

In the remainder of this section, we show that the synthesis fragment 
is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
Following our formulation as second order satisfaction, %the framing of the halting problem as a program synthesis problem,
in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.

@@

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Unconditional Termination Formula {\bf [UT]}]
\label{def:UT}
\begin{align*}
 \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

\noindent {\bf Unconditional termination.}
To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
state it starts in), we can find a ranking function for $L$.

\begin{theorem}
\label{thm:ut}
 The loop $L(G, T)$ terminates from every start state iff Definition~\ref{def:UT} is
 satisfiable.
\end{theorem}

@@

The existence of a ranking function is equivalent to the satisfiability
of the formula from Definition~\ref{def:UT}, so a witness to the satisfiability
of formula {\bf [UT]} is also a ranking function and thus a proof of $L$'s unconditional termination.

Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.\\

@@

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
i.e. a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$. 
For our case, given a loop $L = \langle G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.

\begin{theorem}
\label{thm:ont}
 The loop $L(G, T)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
\end{theorem}

@@

\begin{theorem}
\label{thm:cnt}
 If $T$ is deterministic, i.e. each state $x$ has exactly one successor $x'$ such that $T(x, x')$, then
 Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
\end{theorem}

\begin{proof}
 If $T$ is deterministic, then each $x$ has a unique successor $S(x)$, so $\forall x, x' . T(x, x') \leftrightarrow x' = S(x)$.

@@

@@

@@

\subsection{Composing a Loop with its Environment}

>>>>

\section{Motivational Examples} \label{sec:motivation}
- see phase change programs.

Figure~\ref{fig:handletable} illustrates the most common assumptions made by existent termination analyses:
\begin{itemize}
\item[(i)] programs are linear.
\item[(ii)] terminating programs have linear termination arguments.
\item[(iii)] bit-vectors semantics and mathematical integers semantics are equivalent.
\item[(iv)] floats semantics and mathematical reals semantics are equivalent.
\end{itemize}

@@

In the rest of this section, we show that the synthesis fragment 
is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
Following our formulation as second order satisfaction, %the framing of the halting problem as a program synthesis problem,
in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.

@@

\begin{figure*}
\begin{framed}
\begin{definition}[Unconditional Termination Formula {\bf [UT]}]
\label{def:UT}
\begin{align*}
 \exists R . \forall x, x' . & G(x) \wedge T_o(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

%% \begin{definition}[Non-deterministic Non-Termination Formula - Closed Recurrence Set {\bf [Nondet-CNT]}]
%% \label{def:nondet-cnt}
%%  \begin{align*}
%%   \exists N, x_0 . \forall x, x' . \exists y_0. & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
%% 							& N(x) \wedge T_u(x, x') \wedge A(x') \rightarrow N(x')~
%%  \end{align*}
%% \end{definition}

@@

\noindent {\bf Unconditional termination.}
To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
state it starts in), we try to find a ranking function for it. When defining $L$'s termination 
formula, we need only consider the over-approximation of its transition relation $T$, which we call $T_o$.
Note that this distinction is only important if $L$'s body contains nested loops, for which we'll have to 
consider the over-approximation of their transition relation's transitive closure
as shown in Section~\ref{sec:env}. For any other statements, $T_o$ and $T$ are equivalent.

\begin{theorem}
\label{thm:ut}
 The loop $L(G, T_o)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}) is
 satisfiable.
\end{theorem}

@@

As the existence of a ranking function is equivalent to the satisfiability
of the formula {\bf [UT]}, a satisfiality witness 
is also a ranking function and thus a proof of $L$'s unconditional termination.

Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.

Note that different formulations for unconditional termination are possible. We are aware of a proof rule based on 
transition invariants, i.e. supersets of the transition relation's transitive closure \cite{DBLP:conf/pldi/GrebenshchikovLPR12}.
This formulation assumes that the second order logic has a primitive predicate for disjunctive well-foundedness. 
Conversely, our formulation in Definition~\ref{def:UT} does not require a transitive closure operator, nor a disjunctive well-foundedness predicate. \\

@@

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
i.e. a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$.
For our case, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition relation, $T_u$. 
%$L = \langle G,T \rangle$
Again, this distinction is mostly important when dealing with nested loops, as illustrated in Section~\ref{sec:env}.

\begin{theorem}
\label{thm:ont}
 The loop $L(G, T_u)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
\end{theorem}

@@

\begin{theorem}
\label{thm:cnt}
 If $T_u$ is deterministic, i.e. each state $x$ for which the guard $G(x)$ holds has exactly one successor $x'$ such that $T_u(x, x')$, then
 Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
\end{theorem}

\begin{proof}
 If $T_u$ is deterministic, then each $x$ satisfying $G(x)$ has a unique successor $S(x)$, so $\forall x, x' . G(x) \wedge T_u(x, x') \leftrightarrow x' = S(x)$.

@@

Note that by defining $T_u$'s determinism as the existence of exactly one successor for any state satisfying $L$'s guard, 
we eliminate the possibility of $T_u$ being $false$, which would otherwise constitute a counter-example for Theorem~\ref{thm:cnt}.

@@

\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
To understand the issue, imagine the following loop with an non-deterministic assignment:

Only for certain values of \texttt{y} the loop is non-terminating.
In order to restrict the values that may be assigned to \texttt{y}, we need an assumption associated with the non-deterministic choice. 
However, such an assumption may cause a reachable state at the non-deterministic assignment to not have any successor. In such a case
the termination formula would be satisfiable, although the execution halts. This situation can be avoided by checking the existence of 
at least one successor for each reachable state at the non-deterministic assignment. Although we did put a reasonable amount of effort into it, 
we failed to come up with a formulation for the non-termination of such non-deterministic transition relations 
inside the synthesis fragment.
%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.

@@

\subsection{Composing a Loop with its Environment} \label{sec:env}

synthesizing environment abstractions: underapp for termination and underapp for non-termination.

\cite{} addresses the problem of automatically underapproximating weakest preconditions.
